---
title: "Exploratory Data Analysis"
author: "Nathan W. Van Bibber"
date: "3/19/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Week 1 

##

## Graphics Devices

Two ways to create a plot:  
1. Call a plotting function, and the plot appears on the screen  
   + Annotate the plot if necessary  
2. For sending a plot to a file device (plot will not actually appear on screen):  
   A. Explicitly launch graphics device  
   B. Call a plotting function to make a plot (copy paste from script)  
   C. Annotate if necessary  
   D. **Important!** Explicitly *close* graphics device with `dev.off()`  

```{r}
library(datasets)
with(faithful, plot(eruptions, waiting)) ## make plot appear on the screen
title(main = "Old Faithful Geyser data")  ## Annotate with a title
```


Two basic types of file  
Vector  
* pdf  
* svg  
* win.meta  

Bitmap  
* png  
* jpeg  
* tiff  
* bmp  



# Week 2

## Lattice Plotting System

Useful for plotting high dimensional data and for making many plots at once.  
* have to create entire plot at once (cannot annotate later)  
* uses *formula* notation eg. `xyplot(y ~ x | f * g, data)`  
  + "~" means y as a function of x  
  + f and g are *conditioning variables* (optional)  
    - "*" indicates an interaction between two variables  
* the second argument is the data frame or list where the variables are found  
  + if missing it will look in your workspace for the variables    
The Lattice plotting system is implemented using the following packages:  
*lattice* - contains code for producing Trellis graphics (independent of "base" graphics) and contains functions like:  
* `xyplot` scatterplots  
* `bwplot` box-and-whiskers plots ("boxplots")  
* `histogram`  
* `stripplot` like a boxplot but with actual points  
* `dotplot` plot dots on "violin strings"  
* `splom` scatterplot matrix (like `pairs` in base plotting system)  
* `levelplot`. `contourplot` for plotting "image" data  
*grid* - the *lattice* package builds ontop of *grid* (implements a different graphing system independent of "base" system)  

```{r}
library(lattice)
library(datasets)
## simple scatterplot
xyplot(Ozone ~ Wind, data = airquality)
```

END