---
title: "ReproducibleResearch"
author: "Nathan W. Van Bibber"
date: "3/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Week 1  

## Concepts and Ideas  
It is difficult to reproduce entire studies, but can't do nothing.
Reproducibility bridges this gap.  

* validation of data analysis
  + complex statistical methods subjected to intense scrutiny

# Week 2

## Coding Standards in R

1. Write code in a text editor and save as a text file
2. Indent your code (4-8 spaces)
3. Limit the width of your code (80 columns)
4. Limit length of functions (ideally fit entire function on one screen)
    * one function for one basic activity
    * easier when using debugger, traceback, or profiler
    
## Markdown

Markdown is a text-to-HTML tool for web writers.  

* Allows you to write in plain text format that is easy to read and write
* then converts to srtucturally valid XHTML (or HTML)  

It is a simplified version of a "markup" language that allows the writer to 
focus on writing and not formatting. It is then easily converted into HTML and other formats
    
*This text is in Italics*  
**this text is Bold**

- list item 1
- list item 2

#### Insert links with format "[Text](url)"
[My GitHub](https://github.com/vanbibn/IntroR-course)  
[My Blog](https://r-youready.blogspot.com/2020/04/assignment-12.html)

#### Alternatively, for a more readable format links can be done like this:
###### Note: I'm not sure if this is still a valid way to do hyperlinks...

~~A book on Data Science I want to read is  [R for Data Science][1] by Hadley Wickham and Garrett Grolemund. There is also [Introduction to Data Science][2] by Rafael Irizarry. Jenny Bryan has a book on using version control called [Happy Git and GitHub for the useR][3]

This format makes your document cleaner and easier to read.

Then, at the bottom of the page put the actual urls.
[1]:https://r4ds.had.co.nz/index.html "R for Data Science"
[2]:https://rafalab.github.io/dsbook/ "Introduction to Data Science"
[3]:https://happygitwithr.com/index.html "Happy Git and GitHub for the useR"~~





#### New lines require a double space at the end of a line  
Is this text all -
on one line?

This text sould be  
On 2 lines!



## R Markdown

R Mrakdown allows for integration of R code 
    * R code is evalusted as part of the processing of the markdown
    * Results from R code are also inserted into the document
    
#### In line Text computations

```{r computetime,echo=FALSE}
time <- format(Sys.time(), "%a %b %d, %Y %X")
```


The current time is `r time`.  
My favorite random number is `r rnorm(1)`.

### Incorporate graphics

First I need to simulate data.
```{r simulatedata,echo=TRUE}
x <- rnorm(100)
y <- x + rnorm(100, sd = 0.5)
```

Then I can create a scatterplot of the data.
```{r scatterplot,fig.height=4}
par(mar = c(5,4,1,1), las = 1)
plot(x, y, main = "My Simulated Data")
```

